---
title: "Modern Statistical Computing"
subtitle: "6. Simulation-based inference"
author: "David Rossell"
institute: "Pompeu Fabra University"
execute:
  echo: true
format:
  revealjs:
    theme: [default, custom.scss]
    scrollable: true
toc: true
toc-depth: 1
number-sections: false
mouse-wheel: true
code-overflow: scroll
code-line-numbers: false
code-copy: true
cache: true
title-slide-style: pandoc
bibliography: references.bib
---

## Reproducing these lecture notes

We load the required R packages

```{r, warning=FALSE}
library(tidyverse)
library(boot)
library(coin)
library(fivethirtyeight)
library(lmPerm)
library(ggpubr)
library(plyr)
```

We also source routines.R, which has auxiliary functions

```{r}
source('../code/routines.R')
```


##

Standard theory to obtain confidence intervals / P-values relies on assumptions

- As $n \rightarrow \infty$ some assumptions are less relevant (e.g. error normality)

- Other assumptions remain critical

**Example.** Let $\hat{\beta}= (X^T X)^{-1} X^T y$, where $y \in \mathbb{R}^n$, then
$$
\begin{aligned}
&E(\hat{\beta})= (X^T X)^{-1} X^T E(y) \\
&\mbox{Cov}(\hat{\beta})= (X^T X)^{-1} X^T \mbox{Cov}(y) X (X^T X)^{-1}
\end{aligned}
$$

If we also assume that $y= X \beta + \epsilon$, with $\mbox{Cov}(y)= \mbox{Cov}(\epsilon)= \sigma^2 I$, then
$$
E(\hat{\beta})= (X^T X)^{-1} X^T X \beta= \beta; \mbox{Cov}(\hat{\beta})= \sigma^2 (X^T X)^{-1}
$$

We assumed **linearity** and **constant variance**. Else the result doesn't hold

# Bootstrap

## Reminder: sampling distribution

Let $\theta$ be a parameter of interest, and $\hat{\theta}=\hat{\theta}(y)$ an estimator that depends on the observed data $y$.
Since $y$ is random, so is $\hat{\theta}$.

**Def.** The sampling distribution of $\hat{\theta}$ is its probability distribution under repeated sampling of $y$


The bootstrap (Efron, 1977) is a method to approximate the distribution of many estimators, under mild assumptions


## Bootstrap

> The sub-sample is to the sample, as the sample is to the population

Let $(y,X)$ be an $n \times (p+1)$ dataset. 
For $b=1,\ldots,B$ do

1. Sample $n$ rows with replacement from $(y,X)$, store into $(\tilde{y}^{(b)},\tilde{X}^{(b)})$

2. Obtain estimator $\tilde{\theta}^{(b)}$ from $(\tilde{y}^{(b)},\tilde{X}^{(b)})$

For large $b$, the distribution of $\tilde{\theta}^{(b)}$ approximates that of $\hat{\theta}$

Note: $n$ cannot be too small, and certain technical conditions on $\hat{\theta}$ are needed

See some [animations](https://www.stat.auckland.ac.nz/~wild/BootAnim/index.html)

## Example. Baseball

Goal: estimate batting average in major league baseball games $\geq$ 1990

- ab: number of at bats

- h: number of hits

```{r}
baseball= baseball[baseball$year >= 1990, ]
head(baseball)
```

---

Batting average is $\theta= E(h) / E(\mbox{ab})$. Consider the estimator
$$
\hat{\theta}= \frac{\sum_{i=1}^n h_i}{\sum_{i=1}^n \mbox{ab}_i}  
$$

Central Limit Theorem: numerator and denominator in $\hat{\theta}$ approx. Normal as $n \rightarrow \infty$. But $\hat{\theta}$ may not be Normal, and anyway what is it's variance? 


## Example. Baseball


Define a function to compute the batting average

```{r}
bat.avg= function(data, indices=1:nrow(data)) {
 sum(data[indices, "h"], na.rm=TRUE) /
 sum(data[indices, "ab"], na.rm=TRUE)
}
bat.avg(baseball)
```

For each boostrapped dataset, compute the batting average

```{r}
set.seed(12345)
B= 2000; n= nrow(baseball)
th= double(B)
for (b in 1:B) {
  idx= sample(1:n, size=n, replace=TRUE)
  th[b]= bat.avg(baseball, indices=idx)
}
```

## Example. Baseball

```{r}
quantile(th, probs=c(.025,0.975)) #Boostrap-based 95% CI
```

:::panel-tabset

### Histogram

```{r}
th= tibble(th)
ggplot(th, aes(x=th)) + geom_histogram(aes(y= ..density..)) + stat_overlay_normal_density(linetype='dashed')
```

### qq-plot

```{r}
ggplot(th, aes(sample=th)) + geom_qq() + geom_qq_line(col='red')
```

:::

## R functions

The beauty is that the bootstrap works in more complicated settings

Function `boot` (package `boot`) has a generic implementation. It requires:

- `data`: the dataset

- `statistic`: function to obtain $\hat{\theta}$. Its 1st argument is the data, its 2nd argument the row indices, such as function `bat.avg` defined earlier

- `R`: number of bootstrap samples

Further options, including parallel computing



```{r}
avgBoot= boot(data=baseball, statistic=bat.avg, R=2000)
summary(avgBoot)
quantile(avgBoot$t, probs=c(0.025,0.975))
```


## Regression example


```{r}
data(mtcars)
dim(mtcars)
```


```{r}
f= formula(mpg ~ cyl + disp + hp + drat + wt + qsec + vs + am)
fit= lm(f, data=mtcars)
coefSummary(fit)
```

These intervals are valid if linear model assumptions hold (linearity, constant variance, uncorrelated errors, large $n$ / normal errors)

## Check model assumptions

```{r}
mtcars.pred= tibble(mpg=mtcars$mpg, mpg_pred= predict(fit))
mtcars.pred$res= mtcars.pred$mpg - mtcars.pred$mpg_pred
```

:::panel-tabset

### Linearity

```{r}
ggplot(mtcars.pred, aes(x=mpg_pred, y=res)) +
  geom_point() +
  geom_smooth()
```

### Error normality

```{r}
ggplot(mtcars.pred, aes(sample=res)) + 
  geom_qq() + 
  geom_qq_line() +
  labs(x='Sample quantile',y='Normal quantile')
```

:::

## Bootstrapped intervals

First, define a function to obtain $\hat{\theta}$

```{r}
bhat= function(data, indices=1:nrow(data), formula) {
  fit= lm(formula, data=data[indices,])
  return(coef(fit))
}
round(bhat(mtcars, formula=f), 3)
```

Second, run `boot`. The bootstrap samples are in element `$t`

```{r}
bhat.boot= boot(mtcars, statistic=bhat, R=2000, formula=f)
dim(bhat.boot$t)
round(bhat.boot$t,3)[1:5,] #show first five bootstrap samples
```

## Sampling distribution of $\hat{\theta}$

Format the data into a long tibble, so we can use faceting

```{r}
colnames(bhat.boot$t)= names(coef(fit))
bhat.boot= as_tibble(bhat.boot$t)
bhat.boot
```

```{r}
bhat.bootlong= pivot_longer(bhat.boot, cols= names(bhat.boot), names_to="variable", values_to="bootstrap_value")
bhat.bootlong
```

## Sampling distribution of $\hat{\theta}$

There are deviations from normality.
Set `scales="free"` so that each panel uses different axis limits

```{r}
ggplot(bhat.bootlong, aes(x=bootstrap_value)) +
  geom_histogram(aes(y= ..density..)) +
  stat_overlay_normal_density(linetype='dashed') +
  facet_wrap(~ variable, scales="free")
```


## Bootstrapped intervals


Bootstrap 95% confidence intervals

```{r}
#map_df gets quantiles for all columns, returns a data frame (tibble)
bhat.ci= map_df(bhat.boot, quantile, probs=c(0.025,0.975), na.rm=TRUE)
bhat.ci= cbind(varname= names(bhat.boot), bhat.ci) 
```

Those from least-squares theory. Some are quite different!

```{r}
coefSummary(fit)
```

---

Let's plot the intervals for easier comparison

```{r}
#| code-fold: true
yvals= 1:length(coef(fit))
ci= cbind(bhat.ci, confint(fit), y.ols=yvals, y.boot=yvals+.1)
names(ci)= c('varname','low.boot','high.boot','low.ols','high.ols','y.ols','y.boot')
ggplot(ci) + 
  geom_segment(aes(x=low.ols,xend=high.ols,y=y.ols,yend=y.ols)) +
  geom_segment(aes(x=low.boot,xend=high.boot,y=y.boot,yend=y.boot), color='red') +
  geom_text(aes(x=low.ols, y=y.ols, label=varname), nudge_y = 0.3) +
  labs(x='Confidence interval', y='') +
  theme(axis.text.y=element_blank(),  axis.ticks.y=element_blank()) #remove y axis labels 
```


# Permutation tests

## Permutation tests

A strategy to test a null hypothesis without relying on parametric assumptions (normality etc.)

We consider two uses

- Compare a mean/median/proportion across groups

- Assess significance of a regression coefficient


## Group comparisons

Let $y_i$ be outcome, $x_i \in \{1,\ldots,K\}$ a group indicator,
$\mu_k= E(y_i \mid x_i=k)$ the expectation in each group

$$H_0: \mu_1= \mu_2= \ldots = \mu_K$$

General recipe to obtain a P-value

1. Define a test statistic $T$ to measure evidence against $H_0$

2. Compute its value in the observed data, denote by $t$

3. Compare $t$ to the distribution of $T$ if $H_0$ were true

$$
\mbox{P-value}= P(T \geq t \mid H_0)
$$

Idea: if $y_i$ independent of $x_i$ (so $H_0$ true) we can permute $x_i$


----

Formally, $H_0: P(y_i, x_i)= P(y_i) P(x_i)$. Define $T$ that targets our original $H_0$ (mean comparison)

$$
T= \sum_{k=1}^K (\bar{y}_k - \bar{y})^2
$$


**Example.** $n=24$ mice, 12 receive a muscle relaxant and 12 placebo. We measure the time (up to 300 sec) that they run in a mill

## Mouse data

:::panel-tabset

```{r}
data("rotarod", package = "coin")
```

### Data

```{r}
rotarod
```

### Scatter plot

```{r}
ggplot(rotarod, aes(group, time)) + geom_point()
```

:::


## Permutation test

First, define function to compute test statistic

```{r}
tstat= function(data) {
  xbar= aggregate(time ~ group, data=data, FUN=mean)
  ans= sum((xbar$time - mean(xbar$time))^2)
  return(ans)
}
```

Second, obtain test statistic for the observed data

```{r}
tobs= tstat(rotarod)
tobs
```

Finally, obtain test stat under permuted data

```{r}
B= 2000; n= nrow(rotarod)
t0= double(B)
for (b in 1:B) {
  groupperm= rotarod$group[sample(1:n, size=n, replace=FALSE)]
  dataperm= data.frame(time=rotarod$time, group=groupperm)
  t0[b]= tstat(dataperm)
}
```

## Distribution under $H_0$


```{r}
pvalue= mean(t0 >= tobs)
pvalue
hist(t0,main='',xlab='Test statistic'); abline(v= tobs, col='blue')
```

## R package coin

`coin` implements permutation tests (with additional options)

```{r}
independence_test(time ~ group, data=rotarod, teststat="quadratic")
```



## Independence between continuous variables

Permutation tests also apply if $x_i$ is continuous

**Example.** State illiteracy vs. murder rate

```{r}
states= as.data.frame(state.x77)
ggplot(data=states, aes(Illiteracy, Murder)) + geom_point()
```

---

Compute Spearman's correlation (a robust alternative to Pearson's correlation)

```{r}
cor(states$Illiteracy, states$Murder, method='spearman')
```

Test its statistical significance

```{r}
set.seed(1234)
spearman_test(Illiteracy~Murder, data=states)
```

Bootstrap interval

```{r}
r= function(data, indices=1:nrow(data)) cor(data[indices,'Illiteracy'],data[indices,'Murder'],method='spearman')

rboot= boot(states, r, R=2000)
quantile(rboot$t, probs=c(.025,.975))
```


## Multiple regression

Goal: does murder rate (per 100,000 population) depend on population, after accounting for literacy, income, and number of days below freezing temperature?

```{r}
sel= c('Population','Illiteracy','Income','Frost')
states[,sel]= scale(states[,sel])  #standardize to zero mean, unit variance
fit1= lm(Murder~Population + Illiteracy+Income+Frost, data=states)
coefSummary(fit1)
```

---

Package `lmPerm` offers a permutation test for regression

```{r}
fit2= lmp(Murder~Population + Illiteracy+Income+Frost, data=states)
summary(fit2)
```

## Permutations in regression

To assess significance of $x_1$ in regression, it is **incorrect** to simply permute $x_1$

If we only permute $x_1$, the null hypothesis is

$$P(y | x_{1}, x_{2})= P(y \mid x_{2}) \mbox{ and }  P(x_{1}, x_{2})= P(x_{1}) P(x_{2})$$
but we just wanted to test $P(y | x_{1}, x_{})= P(y_i \mid x_{2})$, i.e. $(x_1,x_2)$ are not independent!

Residual permutation (used by `lmPerm`) is a way to address this 


## Residual permutation

Idea: replace $x_{i1}$ by a permuted version that is correlated with $x_{i2}$

1. Regress $x_1$ on $x_2$ by OLS. Let $\hat{x}_{i1}$ be the predictions, $e_i$ the residuals

2. For $b=1,\ldots,B$, do

- Permute $e_i$ into $\tilde{e}_i^{(b)}$

- Let $\tilde{x}_{i1}^{(b)}= \hat{x}_{i1} + \tilde{e}_i^{(b)}$

- Let $\tilde{\beta}^{(b)}$ be regression estimate of $y$ on $(\tilde{x}_1,x_2)$


By construction $\mbox{cor}(\tilde{x}_1,x_2) \approx \mbox{cor}(x_1, x_2)$. But given $\hat{x}_1$ (i.e. given $x_2$), $\tilde{x}_1$ is independent of $y$


---

Define an R function (saved in routines.R)

```{r}
#Residual permutation test for a single variable in linear regression
#Input
# - y: outcome
# - x1: covariate of interest
# - x2: other covariates
# - B: number of residual permutations to consider
# Output: estimated coefficients for x1 in the B permutations
lmResPerm.onevar= function(y, x1, x2, B=5000) {
  if (!is.matrix(x2)) x2= as.matrix(x2)
  fit= lm(x1 ~ x2)
  x1hat= predict(fit)
  e= residuals(fit)
  bperm= double(B)
  for (b in 1:B) {
    eperm= e[sample(1:length(e), size=length(e), replace=FALSE)]
    x1tilde= x1hat + eperm
    fitperm= lm(y ~ x1tilde + x2)
    bperm[b]= coef(fitperm)['x1tilde']    
  }
  return(bperm)  
}
```

---

Obtain a residual permutation P-value  for Population

```{r}
y= states$Murder; x1= states$Population; x2= states[,c('Illiteracy','Income','Frost')]
bperm= lmResPerm.onevar(y, x1, x2, B=5000)
```

```{r}
bobs= coef(fit1)['Population']
pvalue= mean(abs(bperm) > abs(bobs))
pvalue
```

Now for all covariates (`lmResPerm` is in routines.R)

```{r}
x= cbind(x1, x2)
fit1.resperm= lmResPerm(y, x, B=5000)
fit1.resperm$pvalue
```

Recall the results from package `lmPerm` 

```{r}
summary(fit2)$coef
```


## Final remarks

Bootstrap is more popular than permutation: if we have CIs, we can also test the null hypothesis

Example: test $H_0: \theta=0$.

- If a 95% CI for $\theta$ doesn't include 0, then P-value$\leq 0.05$ 

- If a 99% CI doesn't include 0, then P-value$\leq 0.01$. 

Let $(1-\alpha)$% be the largest CI that doesn't include 0. Then P-value$=\alpha$


Dependent data: bootstrap needs to be adjusted (model-based bootstrap, block-based bootstrap...). See package `boot`


# Exercise 

Data `hate_crimes`. Outcome: pre-election number of hate crimes per 100,000 population by the FBI

- education (% adults $\geq$ 25 with high school degree)

- % of people who voted Donald Trump

- Income inequality (Gini index)

In these data the linearity assumption is violated. 

1. Compare the intervals reported by `lm` to those from bootstrap

2. Compare the permutation test P-values to those from `lm`

---

Load packages and format data.

```{r}
library(tidyverse)
library(fivethirtyeight)
source("~/github/statcomp/code/routines.R")
sel= c('state_abbrev','avg_hatecrimes_per_100k_fbi','share_vote_trump','gini_index','share_pop_hs')
hc= hate_crimes[,sel]
names(hc)= c('state','hatecrimes_fbi','votes_trump','gini','hs')
hc= filter(hc, !is.na(hatecrimes_fbi)) #outcome must be non-missing
```

The model is fit by least-squares as follows.

```{r}
fit= lm(hatecrimes_fbi ~ votes_trump + gini + hs, data=hc)
```


