---
title: "Modern Statistical Computing"
subtitle: "9. Multilevel models"
author: "David Rossell"
institute: "Pompeu Fabra University"
execute:
  echo: true
format:
  revealjs:
    theme: [default, custom.scss]
    scrollable: true
toc: true
toc-depth: 1
number-sections: false
mouse-wheel: true
code-overflow: scroll
code-line-numbers: false
code-copy: true
cache: true
title-slide-style: pandoc
bibliography: references.bib
---

## Reproducing these lecture notes

Load required R packages

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(coefplot)
library(DT)
library(GGally)
library(lme4)
source("../code/routines.R")
```

Disclaimer: these notes borrow heavily from Roback & Legler's [Beyond multiple linear regression](https://bookdown.org/roback/bookdown-BeyondMLR) (free online)


# Motivation

## Multilevel models

We often encounter data that was collected at multiple levels

**Example.** Musician's performance stress (negative affect score, PANAS questionnaire)

Performance covariates (Level 1)

- Type of performance: solo, small ensemble, large ensembl

- Audience: who attended (instructor, public, students, jury)

Musician covariates (Level 2)

- Negative emotionality scale from MPQ questionnaire

- Gender, instrument...

Goal: how does stress depend on emotionality, accounting for other covariates?

Issue: data clearly non-independent


---

`musicdata.txt` has several diary entries for each musician (`id`). In total, 497 performances by 37 musicians

```{r}
#convert id to factor, center mpqnem, create Large ensembl indicator
music= as_tibble(read.table("../datasets/musicdata.txt", header=TRUE)) |>
  mutate(id= factor(id), mpqnem= mpqnem-mean(mpqnem), audience= ifelse(perform_type=='Large Ensemble', "large", "small")) |>
  relocate(id, diary, na, mpqnem, audience)
music
```


## Other examples

Two-level models used in many examples

- Housing prices vs. house & neighborhood covariates

- Ambiguity of political tweets made in various districts vs. politician & district covariates

One can also consider >2 levels, for example

- Student marks vs. student, class & school covariates

- Football player's salary vs. player, team & league covariates

We focus on 2 levels for simplicty


## Music data

We focus on 

- Negative affect score (outcome): `na`

- Audience size (large vs. small/solo) (Level 1 covariate): `Large`

- Negative emotionality (Level 2 covariate): `mpqnem`


---

:::panel-tabset

### Scatter-plot

We can explore data as usual

```{r}
ggplot(music, aes(mpqnem, na, color=audience)) +
  geom_point() +
  geom_smooth() +
  labs(x='Negative emotionality', y="Stress (negative affect score)")
```

### Musician id

But recall that data are dependent within a musician

```{r}
ggplot(music, aes(mpqnem, na, color=id)) +
  geom_point() +
  geom_smooth() +
  labs(x='Negative emotionality', y="Stress (negative affect score)")
```

:::


## Option 1. Standard linear regression

The analysis below is wrong. It assumes that all observations are independent, giving too narrow CIs and too small P-values (inflated false positives)

```{r}
fit1= lm(na ~ mpqnem + audience, data=music)
coefSummary(fit1)
```

## Option 2. Two-stage analysis

Consider fitting a model for each musician at level 1: stress vs. performance type. 
Note: we can only use musicians with the 3 performance types


```{r}
#| code-fold: true

fitmodel= function(df) {   #Return OLS
  naudience= length(unique(df$audience))
  if (naudience == 2) {
    ans= round(coef(lm(na ~ audience, data=df)),3)
  } else {
    ans= rep(NA, 2)
  }
  names(ans)= c('Intercept','small')
  return(ans)
}

music_nest= group_by(music, id) |> nest() 
b1= tibble(id=music_nest$id, map_df(music_nest$data, fitmodel))
DT::datatable(b1)

```



---

Step 1 coefficients are interpreted as usual

- Intercept: average musician stress (for reference category)

- Small ensembl/Solo: increased stress for small/solo peformance

Step 2: regress estimates from Step 1 on level 2 covariate (negative emotionality)

```{r}
b1= unique(select(music, id, mpqnem)) |>
  left_join(b1, by='id')
b1
```


---


:::panel-tabset

### Intercept

```{r}
coefSummary(lm(Intercept ~ mpqnem, data=b1))
```

```{r}
ggplot(b1, aes(mpqnem, Intercept)) + geom_point() + geom_smooth()
```

### Audience

```{r, warning=FALSE}
coefSummary(lm(small ~ mpqnem, data=b1))
```

```{r, warning=FALSE}
ggplot(b1, aes(mpqnem, small)) + geom_point() + geom_smooth()
```

:::

## Issues

```{r}
nmissing= sum(is.na(b1$Intercept))
```

The two-step approach suffers from several issues

- Estimates unavailable for some musicians (`r nmissing` in our example)

- All musicians receive the same weight, regardless of how much data (performances) we have (estimates precision)

- Suboptimal way to share info across musicians

---

Further subtle but relevant issue $(\hat{\beta}_{1i},\hat{\beta}_{2i},\hat{\beta}_{3i})$ for musician $i$ are correlated: using three separate regressions not ideal

```{r, warning=FALSE}
#| code-fold: true
ggpairs(b1[,c('Intercept','small')])
```

# A proper multi-level model

---

Let $y_{ij}$ be stress for subject $i$ in performance $j$

**Level 1:** $$y_{ij}=  \beta_{0i} + \beta_{1i} \mbox{small}_{ij} + \epsilon_i$$

**Level 2:**
$$
\begin{aligned}
\beta_{0i}&= \alpha_0 + \alpha_1 \mbox{mpqnem}_i + u_i
\\
\beta_{1i}&= \gamma_0 + v_i
\end{aligned}
$$
where $\epsilon_i \sim N(0,\sigma^2)$ and $(u_i,v_i) \sim N(0, \Sigma)$
$$
\Sigma= \begin{pmatrix} \sigma_u^2 & \rho \sigma_u \sigma_v \\ \rho \sigma_u \sigma_v & \sigma_v^2 \end{pmatrix}
$$
As discussed, the intercept $\beta_{0i}$ and slope $\beta_{1i}$ may be correlated. Note: one could add mpqnem to the $\beta_{1i}$ regression 


## Composite model

Re-write the model, plugging in the expressions for $\beta_{0i}$ and $\beta_{1i}$ into Level 1

$$
\begin{aligned}
&y_{ij}=  \beta_{0i} + \beta_{1i} \mbox{small}_{ij} + \epsilon_i=
\\
&\alpha_0 + \alpha_1 \mbox{mpqnem}_i + u_i +
(\gamma_0 + v_i) \mbox{small}_{ij} + \epsilon_i=
\\
&\alpha_0 + \alpha_1 \mbox{mpqnem}_i + \gamma_0 \mbox{small}_{ij}
+ u_i + v_i \mbox{small}_{ij} + \epsilon_i
\end{aligned}
$$

- Main effects for mpqnem and small

- Random intercept for individual $i$ is $\alpha_0 + u_i$, random slope is $\gamma_0 + v_i$

Key points

1. If $\alpha_1=0$, mpqnem has no effect (our primary goal)

2. $(u_i,v_i)$ imply dependent observations within musician $i$


---

The model can be fit with `lmer` (package `lme4`). The formula specifies the composite model (small audience nested within musician)

```{r}
fit2= lmer(na ~ mpqnem + audience + (audience | id), REML=TRUE, data=music)
summary(fit2)
```


---

Estimates similar to least-squares, but CIs usually wider

```{r}
multiplot(fit1, fit2, intercept=FALSE, names=c("Ordinary least-squares","Multilevel model"))
```


---

We can extract inference as usual

:::panel-tabset

### Fixed effects

```{r}
summary(fit2)$coef
```

### Conf. int

```{r}
confint(fit2)
```

## Random effects

```{r}
coef(fit2)
```

:::








# Statistical inference

## Parameter estimation

$$
y_{ij}= \alpha_0 + \alpha_1 \mbox{mpqnem}_i + \gamma_0 \mbox{small}_{ij}
+ u_i + v_i \mbox{small}_{ij} + \epsilon_i \mbox{, } (\epsilon_i, u_i, v_i) \sim N(0,\Sigma)
$$

- Fixed effects: $\alpha_0, \alpha_1, \gamma_0$, variance components: $\Sigma$

- Random effects: $u=(u_1,\ldots,u_n)$, $v=(v_1,\ldots,v_n)$


Two popular criteria to fit multilevel models

- MLE: $\max p(y \mid \alpha_0, \alpha_1, \gamma_0, \Sigma, u, v)$

- Restricted MLE: first $\max_{\alpha, \gamma, \Sigma} p(y \mid \alpha_0, \alpha_1, \gamma_0, \Sigma)$, then
$$
\max_{u,v} p(y \mid \hat{\alpha}_0, \hat{\alpha}_1, \hat{\gamma}_0, \hat{\Sigma}, u, v)
$$

## MLE vs RMLE

MLE a bit worse than RMLE (default in `lmer`)

- Larger estimation error for fixed effects

- Biased for variance components (under-estimation)

- But, MLE allows testing the removal of fixed effects via likelihood-ratio test

Testing the removal of random effects is a bit more complex



## Testing fixed effects

As with GLMs, `anova` tests whether fixed effects can be removed (it uses MLE internally).
We also get AIC and BIC.

```{r, eval=FALSE}
fit2= lmer(na ~ mpqnem + audience + (audience | id), REML=TRUE, data=music)
```

```{r}
fit3= lmer(na ~ audience + (audience | id), REML=TRUE, data=music)
anova(fit2, fit3)
```




## Testing variance components

Since random slopes for audience have small variance, consider a model where only the intercept is random
(same effect for small audience across individuals)

:::panel-tabset

### Random intercept & slope

```{r, eval=FALSE}
fit2= lmer(na ~ mpqnem + audience + (audience | id), REML=TRUE, data=music)
```

```{r}
summary(fit2)
```

### Random intercept

```{r}
fit4= lmer(na ~ mpqnem + audience + (1 | id), REML=TRUE, data=music)
summary(fit4)
```

:::


## Testing variance components

Random slopes $v_i \sim N(0, \sigma_v^2)$, test the null hypothesis $H_0: \sigma_v^2$

Likelihood-ratio test has a complex asymptotic distribution (not $\chi^2$)

Better to use a "parametric Boostrap test"

1. Fit the null model (e.g. no random slopes). Use it to simulate many datasets

2. For each dataset, fit the larger model & compute the likelihood-ratio test statistic

3. P-value: proportion of simulated LRTs > observed data LRT

---

Easy to do in R: function `bootstrapAnova`. Below only `B=100` samples are used so that it runs fast, in practice you want at least 1000

```{r}

bootstrapAnova(modelA= fit2, model0= fit4, B=100)

```



What model do the AIC/BIC favor? According to BIC, can we drop the random intercept?


```{r}
fit5= lm(na ~ mpqnem + audience, data=music)

BIC(fit5)
```


# Exercise

---

Consider 4 covariates

- Level 1. `audience` (small/large), `bymemory` (played by memory)

- Level 2. `mpqnem` (negative emotionality score)

Fit a model with fixed effects for all 3 covariates, and random slopes for `audience` and `bymemory`

- Interpret the estimated fixed effect of `bymemory`. What association with stress does it suggest?

- Is `bymemory` significantly associated with stress?

Getting started

```{r, eval=FALSE}
library(tidyverse)
library(lme4)
source("../code/routines.R")

music= as_tibble(read.table("~/github/statcomp/datasets/musicdata.txt", header=TRUE)) |>
  mutate(id= factor(id), mpqnem= mpqnem-mean(mpqnem), audience= ifelse(perform_type=='Large Ensemble', "large", "small"), bymemory= ifelse(memory=='Memory',1,0) ) |>
  relocate(id, diary, na, mpqnem, audience)
```





