---
title: "Modern Statistical Computing"
subtitle: "9. Reports"
author: "David Rossell"
institute: "Pompeu Fabra University"
execute:
  echo: true
format:
  revealjs:
    theme: [default, custom.scss]
    scrollable: true
toc: true
toc-depth: 1
number-sections: false
mouse-wheel: true
code-overflow: scroll
code-line-numbers: false
code-copy: true
cache: true
title-slide-style: pandoc
bibliography: references.bib
---

## Reproducing these lecture notes

Load required R packages

```{r, warning=FALSE}
library(tidyverse)
library(dygraphs)
library(DT)
library(fivethirtyeight)
library(gapminder)
library(gt)
library(maps)
library(plotly)
library(threejs)
library(tidyquant)
library(shiny)
```

## Reporting results

R provides useful formats


- Dashboard: short website displaying your main figures/tables/data 

- htmlwidgets: interactive html documents

- Shiny apps: applications allowing for more user input


We cover a gentle intro. Further resources

- [Quarto documentation](https://quarto.org) and [manual](https://quarto.org/docs/guide)

- Hadley Wickam's chapter on [Quarto formats](https://r4ds.hadley.nz/quarto-formats.html)

- [htmlwidgets website](https://www.htmlwidgets.org/index.html)


# Dashboard

File -> New file -> Quarto document. Select html format

Your document should start like this

```{r, eval=FALSE}
---
title: "Diamonds dashboard"
format: html
execute:
  echo: false
---
```


The `echo: false` option disables the printing of code (only output is displayed)

To display individual code chunks use `#| echo: true`

````{verbatim}

```{r}
#| echo: true
2 * 2
```

````

---

You can create multiple tabs as follows

- Write `:::panel-tabset` at start and `:::` at the end

- Start a new tab with `## Title`. Then put R code


````{verbatim}

:::panel-tabset

## Scatterplot

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_point()
```

## Histogram

```{r}
ggplot(diamonds, aes(price)) +
  geom_histogram()
```

:::

````

The next slide shows the result

---

:::panel-tabset

## Scatterplot

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_point()
```

## Histogram

```{r}
ggplot(diamonds, aes(price)) +
  geom_histogram()
```

:::

## Plot layout

Code chunks have layouts attributes to arrange the output

- `#|layout-nrow`: number of rows

- `#|layout-nrow`: number of columns

- `#| layout: [] `: rows / columns of different sizes

It may also be useful to define captions

- `#| fig-cap`: text for figure caption

- `#| cap-location`: where to place to caption (top, bottom or margin)

- `#| fig-subcap`: figure sub-captions

## Example

2 plots in a single row, with figure captions

````{verbatim}

```{r}
#| layout-nrow: 1
#| fig-cap: 
#|   - "Histogram of prices"
#|   - "Prices vs carats"
ggplot(diamonds, aes(price)) +
  geom_histogram()
ggplot(diamonds, aes(carat, price)) +
  geom_point()
```

````

The result is shown in the next slide


##

```{r, echo=FALSE}
#| layout-nrow: 1
#| fig-cap: 
#|   - "Histogram of prices"
#|   - "Prices vs carats"
ggplot(diamonds, aes(price)) +
  geom_histogram()
ggplot(diamonds, aes(carat, price)) +
  geom_point()
```


---

Custom layouts: specify for each row the size of each figure

Example: 2 figures in 1st row of size 0.5 each, 1 figure  in 2nd row

````{verbatim}

```{r}
#| layout: [[0.5,0.5], [1]]
ggplot(diamonds, aes(carat)) +
  geom_histogram()
ggplot(diamonds, aes(price)) +
  geom_histogram()
ggplot(diamonds, aes(carat, price)) +
  geom_point()
```

````

---

```{r, echo=FALSE}
#| layout: [[0.5,0.5], [1]]
ggplot(diamonds, aes(carat)) +
  geom_histogram()
ggplot(diamonds, aes(price)) +
  geom_histogram()
ggplot(diamonds, aes(carat, price)) +
  geom_point()
```


## Displaying tables & data

- `gt` in package `gt`: styling transformations to display a table (column names, alignment etc.)

- `datatable` in package `DT`: display data as html. Provides filtering, pagination, search & sorting

An example by Hadley Wickam: file `../examples/diamonds_dashboard.qmd`


# htmlwidgets

---

Some packages to create interactive plots without using any Javascript (if you know Javascript, then more advanced options are possible)

- Time series plots: [dygraphs](https://rstudio.github.io/dygraphs)

- Make a ggplot interactive: [plotly](https://plotly.com/r)

- 3D scatter-plots and globes: [threejs](https://github.com/bwlewis/rthreejs)

- Data tables: [DT](https://rstudio.github.io/DT)

## dygraphs

`dygraph` visualizes times series interactively

- 1st argument is the data. Either a time series (`xts` object) or a data.frame where the 1st column is the x-axis

- Other arguments to set labels etc.


**Example.** Female and male deaths from lunch cancer

```{r}

lungDeaths= cbind(mdeaths, fdeaths)
lungDeaths[1:5,]

```

---

Zoom into x-axis regions. Double-click to zoom out

```{r}
dygraph(lungDeaths)
```

## Some plot options

```{r}
dygraph(lungDeaths, main = "Deaths from Lung Disease (UK)") |>
  dyOptions(drawPoints = TRUE, pointSize = 3) |>              #show points
  dyHighlight(highlightSeriesOpts = list(strokeWidth = 3)) |> #thicker lines
  dyRangeSelector()                                           #add zooming panel
```


## Plotly

Plotly dynamically displays info about elements of a gg-plot. See [here](https://plotly.com/r) for a gallery of what's possible

**Example.** Scatter-plot for gapminder data

Save the ggplot into an object, then use `ggplotly`. Argument `tooltip`: what aesthetics from your ggplot (**not** from your original dataset) to show

```{r}
#| output-location: slide

library(plotly)

gm2007= filter(gapminder, year==2007)

p= ggplot(gm2007, aes(gdpPercap, lifeExp, text=country)) +
  geom_point(aes(colour=continent, size=pop), alpha = 0.7, show.legend = TRUE) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  labs(x='GDP per capita', y='Life expectancy')

ggplotly(p, tooltip=c("country","gdpPercap","lifeExp"))
```

---

Alternatively, you can use `plot_ly` and `layout` to create plots directly, as illustrated below.


```{r}
fig= plot_ly(gm2007, x = ~continent, y = ~lifeExp, split = ~continent, type = 'violin', box = list(visible = T)) 

layout(fig, xaxis = list(title = "Continent"), yaxis = list(title = "Life Expectancy (years)",zeroline = F))
```


## plotly for time series

```{r}
getSymbols("GOOG", from="2020-01-01", to="2021-12-31") #creates dataset GOOG
```

```{r}
GOOG[1:5,]
```

Compute stock market value relative to start date

```{r}

stock= data.frame(GOOG$GOOG.Adjusted)
stock$GOOG.Adjusted= stock$GOOG.Adjusted/stock$GOOG.Adjusted[1]
stock= data.frame(stock,rownames(stock))
colnames(stock)= c('GOOG','date')
```

```{r}
#| output-location: slide

fig= plot_ly(stock, type = 'scatter', mode = 'lines') |>
  add_trace(x = ~date, y = ~GOOG, name = 'GOOG') |>
  layout(showlegend = F)
options(warn = -1)

layout(fig, xaxis = list(zerolinecolor = '#ffff', zerolinewidth = 2, gridcolor = 'ffff'), yaxis = list(zerolinecolor = '#ffff', zerolinewidth = 2, gridcolor = 'ffff'), plot_bgcolor='#e5ecf6', width = 900)

```

## 3D scatterplots

```{r}

sel= c('state_abbrev','avg_hatecrimes_per_100k_fbi','share_vote_trump','gini_index')
hc= hate_crimes[,sel]
names(hc)= c('state','hatecrimes_fbi','votes_trump','gini')
hc= filter(hc, !is.na(hatecrimes_fbi)) 

labs= c("Votes for Trump","Gini","Hate crimes FBI")
scatterplot3js(hc$votes_trump, hc$gini, hc$hatecrimes_fbi, axisLabels=labs)  |>
points3d(hc$votes_trump, hc$gini, hc$hatecrimes_fbi, color="red", pch=hc$state)

```

## globejs

You can plot data on the globe, e.g. arcs

```{r}
#| code-fold: true
#| colum: screen-inset

f= flights()
f= mutate(f, dest=paste(round(f[,3],2),":",round(f[,4],2),sep="")) # approx location as factor

# select destinations with frequency >300
freq= group_by(f, dest) |> 
  summarise(nflights_dest=n())
fsel= left_join(f, freq) |>
  filter(nflights_dest > 300)

globejs(arcs=fsel[,1:4], arcsLwd= fsel$nflights_dest)

```

## globejs

Or bars, e.g. city populations below

```{r}
#| code-fold: true
#| column: screen-inset

data(world.cities, package="maps")
cities <- world.cities[order(world.cities$pop, decreasing=TRUE)[1:1000],]
value  <- 100 * cities$pop / max(cities$pop)
col <- colorRampPalette(c("cyan", "lightgreen"))(10)[floor(10 * value/100) + 1]
globejs(lat=cities$lat, long=cities$long, value=value, color=col, atmosphere=TRUE)

```


## Data tables

`datatable` from package `DT` displays a dynamic table that can be searched, filtered, sorted... For a full description see [here](https://rstudio.github.io/DT)


```{r}

datatable(gapminder, rownames=FALSE, caption="Life expectancy", options = list(pageLength=5))

```


# Shiny apps


##

Widgets allow *user-based* interactivity: all operations occur in your browser

Shiny apps also allow *server-based* interactions: operations are run by R in a server

To start a Shiny app

1. Create a directory for the app in your computer

2. Add a file `app.R` specifying how the app should look

For a primer on Shiny, see Hadley Wickam's [Mastering Shiny](https://mastering-shiny.org/index.html). See also 

- [Shiny cheatsheet](https://posit.co/wp-content/uploads/2022/10/shiny-1.pdf)

- [Shiny extensions](https://github.com/nanxstats/awesome-shiny-extensions)


## app.R

A minimal example

```{r, eval=FALSE}
library(shiny)

ui= fluidPage(  #specify user interface

    selectInput("color", label="Color", choices= c("Blue","Red","Green")), #dropdown menu
    
    checkboxInput("married", "Are you married?", value=FALSE), #check-box

)

server= function(input, output, session) {  #server calculations
}

shinyApp(ui, server) #run the App
```

Click the "Run App" button in RStudio (or outside RStudio, 
call `shiny::runApp()` with the directory containing `app.R`)


---

In `server` you store all results in `output`, which is a list. The returned element must be the result of a render function:

- `renderPlot`: return a plot (`plotOutput` in `ui`)

- `renderPrint`: print results as in R console (paired with or `verbatimOutput` in `ui`)

- `renderText`: return single string (paired with `textOuput` in `ui`)

- `renderTable`: print a static table (`tableOutput` in `ui`). For short tables

- `renderDataTable`: print a dynamic table (`dataTableOutput` in `ui`). For large tables/data

```{r, eval=FALSE}
ui= fluidPage(  #specify user interface

    selectInput("color", label="Color", choices= c("Blue","Red","Green")), #dropdown menu
    
    verbatimTextOutput("tablehomeworld"), #print output$tablehomeworld

)


server= function(input, output, session) {  #server calculations

  output$tablehomeworld= renderPrint({
    data= filter(starwars, skin_color == input$color) #input$color from selectInput
    table(data$homeworld)
  })
  
}
```

## Some useful functions

- Print text. `p("aaa")`, `strong("bold face text")`, `helpText("italics text")`

- Headers. `titlePanel`, `h1("aaa")`, `h2("aaa")`, `h3("aaa")`

- Line break. `br()`

- Update results button. `submitButton("Get results")`

- Panel arrangement: `sidebarLayout`, `mainPanel`, `sidebarPanel`


## Capturing inputs


```{r, eval=FALSE}
animals= c("dog", "cat", "other") #define vector used multiple times below

ui <- fluidPage(
  selectInput("animal", "Favourite animal?", animals, multiple=FALSE), #single choice
  radioButtons("animal", "Favourite animal?", animals), #single choice
  checkboxGroupInput("animal", "Animals you like?", animals) , #multiple choice

  textInput("name", "What's your name?"), #short text box
  textAreaInput("story", "Tell me about yourself", rows = 3)         #larger text box
  
  numericInput("num", "Number one", value = 0, min = 0, max = 100),  #numeric box
  sliderInput("num2", "Number two", value = 50, min = 0, max = 100), #slider
  
  dateInput("dob", "When were you born?"),                 #single date
  dateRangeInput("holiday", "Departure and return date?")  #date range
  
  fileInput("upload", NULL) #input file
)
```

- 1st argument: identifier of the variable where input is stored

- 2nd argument: text to display

- `value`: default value (optional)

- Other arguments: specific on each input box


## Exercise

Check out `../examples/shinyapp/app.R`. Using this as a template, create a Shiny app where

- The user selects a country in the `gapminder` data (package `gapminder`) 

- A scatterplot of life expectancy vs. year is produced using only the data from that country, adding a linear regression line

- The estimated regression coefficients are reported as separate text output

